HO unification from object language to meta language

### 2024-02-22 paper/main.tex
reduce space before equation
wip

\newcommand{\U}{\ensuremath{=_o}\xspace}
\newcommand{\nU}{\ensuremath{\neq_o}\xspace}
\newcommand{\Ue}{\ensuremath{=_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\neq_\lambda}\xspace}
\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}}
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}

@@

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \U using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi comes equipped with the equational theory \Ue, comprising
  $\eta\beta$ equivalence
  and higher order unification restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \U to feature the same equational theory as \Ue but on the
  object logic DTT. Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \U on top of \Ue for the encoding \Fo.

@@

\noindent
The root cause of the problems we face is that the unification procedure
\Ue of the meta language is not aware of the equational theory of the object
logic \U, even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

@@

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to definition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher order problems
restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
We call this unification procedure \U{}.

@@

In spite of the similarity the link between \Ue{} and \U{} is not trivial,
since the abstraction and application term constructors
the two unification procedure deal with are different. For example

\begin{tabular}{lcl}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[c"f", x]} & \U{} & \elpiIn{c"f"}\\
\elpiIn{lam A x\ app[c"f", x]} & \nUe{} & \elpiIn{c"f"} \\
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \U{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}

@@

To state precisely the problem we solve we need a \Fo{} representation
of DTT terms, a \Ho{} one. Then for any problem $s_1 \U{} s_2$ for
$s_i$ in \Fo{}, we get a  $t_1 \Ue{} t_2$ for $s_i$ in \Fo{} and
a $L = \bigwedge_i \mathrm{link}~ s_i~ t_i $ such that:

\begin{align*}
s_1 \U{} s_2 & \Leftrightarrow  t_1 \Ue{} t_2 \land L \\
s_1 \nU{} s_2 & \Rightarrow t_1 \nUe{} t_2 \lor \lnot L
\end{align*}

where $\mathrm{comp}~t_i = (s_i, \mathrm{link}~ s_i~ t_i)$ and 
link is a predicate allowed to use \U{} on ground terms (hence equal, not unif)

These properties allow us to simulate a unification based
backward search on DTT by using \Ue{}. In particular
any unification that would work in DTT will work,
and whenever a unification in DTT would have failed
either the corresponding \Ue{} or the link fails.

This grants that the trace of the logic program
performing the search not only gives the same result, but
also takes the same paths, that is it fails as early as possible.

>>>>

\newcommand{\EqualRel}{\ensuremath{=}}
\newcommand{\nEqualRel}{\ensuremath{\new}}
\newcommand{\UnifRel}{\ensuremath{\simeq}}
\newcommand{\nUnifRel}{\ensuremath{\not\simeq}}

\newcommand{\Uo}{\ensuremath{\UnifRel_o}\xspace}
\newcommand{\nUo}{\ensuremath{\nUnifRel_o}\xspace}
\newcommand{\Eo}{\ensuremath{\EqualRel_o}\xspace}
\newcommand{\nEo}{\ensuremath{\nEqualRel_o}\xspace}

\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}

\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}} % space non va
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}

@@

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \Uo using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \Uo to be as powerful as \Ue but on the object logic DTT.
  Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \Uo on top of \Ue for the encoding \Fo.

@@

\noindent
The root cause of the problems we face is that the unification procedure
\Ue of the meta language is not aware of the equational theory of the object
logic \Uo, even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

@@

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to definition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher order problems
restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
We call this unification procedure \Uo{}.

@@

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedure deal with are different. For example

\begin{tabular}{lcl}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[c"f", x]} & \Uo{} & \elpiIn{c"f"}\\
\elpiIn{lam A x\ app[c"f", x]} & \nUe{} & \elpiIn{c"f"} \\
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}

@@

To state precisely the problem we solve we need a \Fo representation
of DTT terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
TODO extend \Eo and \Ee with reflexivity on uvars.

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$ and we assume that the unification of our meta language is correct:

$$
t_1 \Ue t_2 \mapsto \sigma \Rightarrow \sigma t_1 \Ee \sigma t_2
$$

\newcommand{\C}[3]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3)}
\newcommand{\D}[3]{\ensuremath{\langle #1,#2 \rangle^{-1}\mapsto #3}}

We illustrate a compilation $\C{s}{t}{l}$ that
maps a term $s$ in \Fo to a term $t$ in \Ho and a list of links $l$.
The links connect unification variables in \Ho with variables
in \Fo and are used to decompile the assignment,
$\D{\sigma}{l}{\rho}$.

Given 
$$
\C{s_1}{t_1}{l_1} \land \C{s_2}{t_2}{l_2}
$$
we define
$$
s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}
    t_1 \Ue t_2 \mapsto \sigma \land
    \D{\sigma}{l_1 + l_2}{\rho}
$$

We write $s \in \mathcal{L}_\lambda$ if all unif variables in
$s$ are applied to distinct bound variables.

\begin{align}
t_1 \Uo t_2 \mapsto \rho &\Rightarrow \rho t_1 \Eo \rho t_2\\
s_i \in \mathcal{L}_\lambda \land \exists \rho, \rho s_1 \Eo \rho s_2 &\Leftrightarrow s_1 \Uo s_2 \mapsto \rho' \subseteq \rho \\
\forall \rho, \rho s_1 \Eo \rho s_2 &\Rightarrow \rho t_1 \Uo \rho t_2
\end{align}

the first one is that we are correct

the second one is that we are complete and mgu iff the problems are nice

the third one is still broken since we did not define how
to apply a fo subst to a ho term but it morally means
that any valid substitution for s1 s2 cannot break
the unification of terms t1 t2 obtained by compiling
s1 and s2 before knowing the subst.

These properties allow us to simulate a unification based
backward search on DTT by using \Ue, in a faithful way:
the trace of the logic program
performing the search not only gives the same result, but
also takes the same paths, that is it fails as early as possible.
